---
title: "Weather and Corn Yield Regressions"
author: "Nathan Mueller, Assignment: Amanda Hastings"
date: "2/25/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download}

# set our API key with NASS
nassqs_auth(key = "AA4F0FBD-6240-3D5A-902F-54E62A4E05F0")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

#### **Linear time trend analysis within Winneshiek County, Iowa**

#### *SLR - Yield by Covariate: Year*

```{r}

library(ggthemes)
library(broom)


#Extract Winneshiek County corn yields 

winne_yields <- cornyields %>%
  filter(county_ansi == 191)

# Fit a model for yield as the response (y) and year as predictor (x) 

winne_timeLM <- lm(yield ~ year, data = winne_yields)
tidy(summary(winne_timeLM))

#Add fitted values from linear model to data frame 

winne_yields$fitted_l <- winne_timeLM$fitted.values

#Plot corn yield(y) by year(x)
#Add line for linear time trend 

ggplot(aes(x = year, y = yield), data = winne_yields) +
  geom_point(shape = 1) +
  geom_line(mapping = aes(x = year, y = fitted_l)) +
  theme_few() +
  labs(x = "Year", y = "Yield (Bu/Acre)", title = "Winneshiek County, IA - Corn Yield (Linear Time Trend)") 

```

With a p-value of 1.77e-13 and less than $\alpha$ = 0.05, we reject the null hypothesis $\H_0 = \beta_1$. We have enough evidence to suggest there is a significant time trend. There estimated 2.457 bu/acre increase in corn yield with each increasing year. 


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

#### **Quadratic time trend analysis within Winneshiek County, Iowa**

#### *MLR - Yield by Covariates: Year,Year^2*

```{r}

# Create column for year^2 

winne_yields$yearsq <- winne_yields$year^2

#Fit quadratic time trend with yield(y) and year and year^2 as covariates(x)

quad_timeLM <- lm(yield~year + yearsq, data = winne_yields)
tidy(summary(quad_timeLM))

#Add fitted values from the quad model to data frame 

winne_yields$fitted_q <- quad_timeLM$fitted.values

#Plot yield(y) by year (x) 
#Add line for quadratic time trend

ggplot(winne_yields) +
  geom_point(mapping = aes(x = year, y = yield), shape = 1) +
  geom_line(mapping = aes(x = year, y = fitted_q)) +
  theme_few() +
  labs(x = expression("Year"), y = "Yield (Bu/Acre)", title = "Winneshiek County, IA - Corn Yield (Quadratic Time Trend)")

```


```{r}

#Plot time trend lines combined 

ggplot(winne_yields) +
  geom_point(mapping = aes(x = year, y = yield), shape = 1) +
  geom_line(mapping = aes(x = year, y = fitted_l, col="Linear Time Trend")) +
  geom_line(mapping = aes(x = year, y = fitted_q, col="Quadratic Time Trend")) +
  theme_few() +
  labs(colour = NULL) +
  labs(x = expression("Year"), y = "Yield (Bu/Acre)", title = "Winneshiek County, IA - Corn Yield Over Time")

```


```{r include=FALSE}

#Plot yield(y) by year^2 (x), use geom_smooth for quadratic line 
# 
# ggplot(aes(x = yearsq, y = yield), data = winne_yields) +
#   geom_point(shape = 1) +
#   theme_few() +
#   labs(x = expression("Year" ^ 2),
#        y = "Corn Yields",
#        title = "Winneshiek County Corn Yield") +
#   geom_smooth(
#     method = "lm",
#     formula = y ~ poly(x, 2),
#     se = FALSE,
#     size = 0.5
#     ,
#     color = "black"
#   )
```


### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

#### **Time series analysis within Winneshiek County, Iowa**

#### *SLR - Yield by Covariate: Average Summer Maximum Temperature*

```{r}

#Add average summer Tmax to winne_yield data 

winne_summer_yield <- left_join(winne_yields,winnesummer, by = "year")%>% 
  filter(!is.na(meantmax))

#Fit a model for yield(y) and average summer maximum temperature(x) alone 

winne_tempLM <- lm(yield~meantmax, data=winne_summer_yield)
tidy(summary(winne_tempLM))

#Add fitted values to data frame 

winne_summer_yield$fitted_2 <- winne_tempLM$fitted.values

```


#### *MLR - Yield by Covariates: Average Summer Maximum Temperature, Year* 

```{r}
#Add summer Tmax^2 to winne_summer_yield 

winne_summer_yield$meantmaxsq <- winne_summer_yield$meantmax^2

#Fit a model for yield(y) with covariates(x): average summer max temp and year 

temp_yearLM <- lm(yield~meantmax + year, data=winne_summer_yield)
tidy(summary(temp_yearLM))

#Add fitted values from model to data frame 

winne_summer_yield$fitted_2a <- temp_yearLM$fitted.values

```


#### *MLR - Yield by Covariates: Average Summer Max Temperature, Max Temperature^2*

```{r}

#Fit a model for yield with covariates: average summer max temp and max temp^2 

quad_tempLM <- lm(yield~meantmax + meantmaxsq, data=winne_summer_yield)
summary(quad_tempLM)

#Add fitted values from model to data frame 

winne_summer_yield$fitted_2b <- quad_tempLM$fitted.values

```


```{r}

#Plot corn yield (y) by Average Summer Maximum Tempearture(x) with fitted values from regressions 

ggplot(winne_summer_yield) +
  geom_point(mapping = aes(x = meantmax, y = yield), shape = 1) +
  geom_line(mapping = aes(x = meantmax, y = fitted_2, col="SLR")) +
  geom_line(mapping = aes(x = meantmax, y = fitted_2a, col="MLR with year")) +
  geom_line(mapping = aes(x = meantmax, y = fitted_2b, col="MLR with temp^2")) +
  labs(colour = NULL) +
  theme_few() +
  labs(x = expression("Average Summer Maximum Temperature ("*degree*C*")"), y = "Yield (Bu/Acre)", title = "Winneshiek County, IA - Corn Yield")
```


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

#### **Cross-section analysis across all Iowa Counties in 2018**

#### *SLR - Yield by Covariate: Average Summer Max Temperature*

```{r}

#Rename county_ansi to countyfp 

corn_yields <- cornyields %>% 
  rename('countyfp'= county_ansi)

#Save countyfp as factor 

corn_yields$countyfp <- as.factor(corn_yields$countyfp)

#Filter for 2018 yields only 

county2018_yields <- corn_yields %>% 
  filter(year==2018)

#Add average summer maximum temp for 2018 

tmax2018 <- tmaxdf %>% 
  filter(year==2018)%>% 
  group_by(countyfp) %>% 
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax))

#Combine average summer max temperature with county yield data for 2018 

countytmax2018 <- inner_join(county2018_yields, tmax2018, by='countyfp')

#Fit a model with yield (y) and average summer Tmax (x) 

yield_2018LM <- lm(yield~meantmax, data=countytmax2018)
tidy(summary(yield_2018LM))

#Add fitted values from model to data frame 

countytmax2018$fitted <-yield_2018LM$fitted.values

#Plot yield(y) by average summer maximum temperature (x)

ggplot(aes(x = meantmax, y = yield), data = countytmax2018) +
  geom_point(shape = 1) +
  geom_line(mapping = aes(x=meantmax, y=fitted))+
  theme_few() +
  labs(x = expression("Average Summer Maximum Temperature ("*degree*C*")"), y = "Yield (Bu/Acre)", title = "Corn Yields across Iowa Counties")
```


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r}

#Combine county and temperature data 

yearly_co_temp <- tmaxdf %>% 
  group_by(countyfp,year)%>% 
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax))

  
co_yearly_join <- left_join(yearly_co_temp, corn_yields,by= c('countyfp','year'))

#Add Tmax^2 to data frame 

co_yearly_join$meantmaxsq <- co_yearly_join$meantmax^2

#Fit a panel regression

panelLM <- lm(yield~ meantmax+ meantmaxsq+countyfp, data = co_yearly_join)
tidy(summary(panelLM))

co_yearly_join$fit_panel <- panelLM$fitted.values

# Plot actual versus fitted yields 

ggplot(co_yearly_join) +
  geom_point(mapping = aes(x = meantmax, y = yield), shape = 1) +
  theme_few() +
  labs(x = expression("Average Summer Maximum Temperature ("*degree*C*")"), y = "Corn Yield")


```


### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r}
#Download NASS data on soybean yields 

#Parameters for soybean data
params <- list(commodity_desc = "SOYBEAN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```


### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
